---
title: "Observable JS vs Shiny for Interactive Statistical Reports"
subtitle: "A Practical Comparison for Data Analysts and Statisticians"
author: "Data Science Team"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    theme: cosmo

---

## Executive Summary

Interactive data visualization has become essential for modern statistical reporting. Two leading technologies for creating interactive reports within the Quarto ecosystem are **Observable JavaScript (OJS)** and **R Shiny**. This white paper provides a comprehensive comparison to help statisticians and data analysts choose the appropriate technology for their reporting needs.

::: {.callout-note}
## Key Takeaway
**Observable JS** excels for self-contained, client-side reports that need broad distribution. **Shiny** excels when you need R's statistical ecosystem or server-side computation for sensitive data.
:::

## Introduction

### The Evolution of Statistical Reporting

Traditional statistical reports were static documents---PDF or Word files with fixed tables and figures. Modern stakeholders increasingly expect interactive elements that allow them to:

- Filter data by subgroups
- Explore different parameter values
- Drill down into specific observations
- Export customized views

Quarto, the next-generation publishing system from Posit, supports both Observable JS and Shiny as interactivity backends, giving statisticians flexibility in their choice of technology.

### Scope of This Comparison

This paper compares Observable JS and Shiny across dimensions critical to statistical reporting:

1. Technical architecture
2. Statistical computing capabilities
3. Deployment and distribution
4. Performance characteristics
5. Learning curve and development experience
6. Security and compliance considerations

## Technical Architecture

### Observable JS: Client-Side Execution

```{mermaid}
%%| fig-cap: "Observable JS Architecture"
%%| fig-width: 5.5
flowchart LR
    A[Quarto Doc] --> B[Static HTML/JS]
    B --> C[Browser]
    C --> D[Computation]
```

Observable JS compiles to static HTML and JavaScript files. When a user opens the report:

- **All code executes in the browser**
- Data is embedded in the document or fetched via HTTP
- No server infrastructure required after initial hosting
- Reactivity is handled by Observable's runtime

### Shiny: Server-Side Execution

```{mermaid}
%%| fig-cap: "Shiny Architecture"
%%| fig-width: 5.5
flowchart LR
    A[Browser] <--> B[Shiny Server]
    B <--> C[R Session]
    C <--> D[DB/APIs]
```

Shiny maintains a persistent R session on a server:

- **Code executes on the server**
- Only UI updates are sent to the browser
- Requires running server infrastructure
- Full access to R's computing environment

### Architectural Comparison

| Aspect | Observable JS | Shiny |
|--------|--------------|-------|
| Execution | Client (browser) | Server (R) |
| State | Browser memory | Server memory |
| Concurrency | Unlimited | Limited by server |
| Network | Initial load only | Continuous |
| Offline | Yes (once loaded) | No |

## Statistical Computing Capabilities

### R/Shiny: The Statistical Powerhouse

Shiny provides direct access to R's comprehensive statistical ecosystem:

**Advantages:**

- **20,000+ CRAN packages** including specialized methods for:
  - Survival analysis (`survival`, `survminer`)
  - Mixed-effects models (`lme4`, `nlme`)
  - Bayesian inference (`brms`, `rstanarm`)
  - Causal inference (`MatchIt`, `WeightIt`)
  - Meta-analysis (`metafor`, `meta`)

- **Established validation** in regulated industries (pharma, finance)

- **Familiar syntax** for statisticians already using R

- **Heavy computation** can run on powerful servers, not limited by browser

```{r}
#| eval: false
#| echo: true
# Example: Complex statistical model in Shiny
# This runs on the server with full R capabilities
output$model_summary <- renderPrint({
  model <- lme4::lmer(
    outcome ~ treatment * time + (1 | subject),
    data = filtered_data()
  )
  summary(model)
})
```

### Observable JS: Modern Visualization Focus

Observable provides excellent visualization with growing statistical capabilities:

**Advantages:**

- **Observable Plot**: Elegant, grammar-of-graphics style plotting
- **D3.js integration**: Unlimited customization potential
- **Arquero**: dplyr-like data transformation in JavaScript
- **Statistical libraries**: `simple-statistics`, `jstat` for basic analyses

**Limitations:**

- Fewer specialized statistical methods
- Less validation history in regulated environments
- Complex models require custom implementation or API calls

```{ojs}
//| eval: false
//| echo: true
// Example: Basic statistics in Observable
import { mean, standardDeviation, linearRegression } from "simple-statistics"

stats = ({
  mean: mean(data.map(d => d.value)),
  sd: standardDeviation(data.map(d => d.value)),
  regression: linearRegression(data.map(d => [d.x, d.y]))
})
```

### Statistical Capability Matrix

| Capability | Observable JS | Shiny |
|------------|:-------------:|:-----:|
| Descriptive statistics | ✓ | ✓ |
| Linear regression | ✓ | ✓ |
| Generalized linear models | Limited | ✓ |
| Mixed-effects models | ✗ | ✓ |
| Survival analysis | ✗ | ✓ |
| Bayesian inference | ✗ | ✓ |
| Time series (ARIMA, etc.) | Limited | ✓ |
| Machine learning | Limited | ✓ |
| Custom visualizations | ✓✓ | ✓ |

## Deployment and Distribution

### Observable JS: Simplicity and Portability

**Deployment options:**

1. **Static file hosting**: Any web server, S3, GitHub Pages
2. **Email attachment**: Self-contained HTML file
3. **Shared drives**: No server needed
4. **Quarto Pub**: Free hosting from Posit

**Advantages for report distribution:**

- Recipients need only a web browser
- No IT infrastructure required
- Works behind firewalls (no external connections)
- Can be archived as single files
- No ongoing server costs

```bash
# Generate self-contained HTML
quarto render report.qmd --to html

# Result: Single HTML file with embedded data and code
# File size: Typically 1-10 MB depending on data
```

### Shiny: Server Infrastructure Required

**Deployment options:**

1. **shinyapps.io**: Managed hosting (free tier available)
2. **Posit Connect**: Enterprise server
3. **Shiny Server**: Self-hosted (open source or pro)
4. **Docker/Kubernetes**: Container deployment

**Considerations:**

- Requires ongoing server maintenance
- Per-user resource consumption
- Session management for concurrent users
- Authentication integration needed for sensitive reports

```bash
# Deploy to shinyapps.io
rsconnect::deployDoc("report.qmd")

# Or render for Posit Connect
quarto render report.qmd
# Then publish via Connect interface
```

### Deployment Decision Matrix

| Scenario | Choice | Rationale |
|----------|--------|-----------|
| Email to stakeholders | OJS | Self-contained HTML |
| Dashboard, <10 users | Either | Shiny may be overkill |
| 100+ concurrent users | OJS | No server scaling |
| Database queries | Shiny | Server-side access |
| Regulated (21 CFR 11) | Shiny | Audit trail support |
| Conference presentation | OJS | No network needed |
| Client deliverable | OJS | No infrastructure |

## Performance Characteristics

### Data Size Considerations

```{r}
#| label: performance-table
#| echo: false
#| tbl-cap: "Performance by Data Size"

performance_data <- data.frame(
  `Data Size` = c("< 10 MB", "10-100 MB", "100 MB-1 GB", "> 1 GB"),
  `Observable JS` = c("Excellent", "Good", "Poor", "Not feasible"),
  `Shiny` = c("Excellent", "Excellent", "Good", "Feasible"),
  check.names = FALSE
)

knitr::kable(performance_data)
```

### Observable JS Performance Profile

**Strengths:**

- Instant reactivity (no network round-trip)
- Smooth animations and transitions
- Responsive UI updates
- CDN caching for repeat visits

**Limitations:**

- Browser memory constraints (~1-4 GB practical limit)
- Initial load time proportional to data size
- No background computation
- Single-threaded (UI can freeze during computation)

### Shiny Performance Profile

**Strengths:**

- Handle arbitrarily large datasets
- Server-side caching across users
- Database connections (query what you need)
- Parallel computation via R packages

**Limitations:**

- Network latency on each interaction
- Server resources scale with users
- Cold start time for new sessions
- WebSocket connection required

### Optimization Strategies

::: {.panel-tabset}

#### Observable JS

```javascript
// Pre-aggregate data to reduce browser load
aggregated = data
  .groupby("category")
  .rollup({
    count: d => op.count(),
    mean_value: d => op.mean(d.value)
  })

// Use efficient data formats
data = FileAttachment("data.parquet").parquet()
```

#### Shiny

```r
# Use server-side filtering before sending to UI
output$plot <- renderPlot({
  # Only process visible subset
  data %>%
    filter(date >= input$date_range[1],
           date <= input$date_range[2]) %>%
    ggplot(...)
})

# Cache expensive computations
model_result <- reactive({
  expensive_model(data())
}) %>% bindCache(input$params)
```

:::

## Learning Curve and Development Experience

### For R Statisticians

| Aspect | Observable JS | Shiny |
|--------|--------------|-------|
| Language familiarity | New (JavaScript) | Familiar (R) |
| Data manipulation | Learn Arquero | Use dplyr |
| Plotting | Learn Plot/D3 | Use ggplot2 |
| Reactivity model | Automatic | Explicit |
| Debugging | Browser DevTools | RStudio |
| Time to productivity | 2-4 weeks | 1-2 weeks |

### Observable JS Learning Path

1. **Week 1**: JavaScript basics, Observable notebook syntax
2. **Week 2**: Observable Inputs and reactivity
3. **Week 3**: Observable Plot for visualization
4. **Week 4**: Arquero for data transformation, integration patterns

**Resources:**

- [Observable tutorials](https://observablehq.com/tutorials)
- [Quarto OJS documentation](https://quarto.org/docs/interactive/ojs/)

### Shiny Learning Path

1. **Week 1**: Shiny UI components and layout
2. **Week 2**: Reactive expressions and observers

**Resources:**

- [Mastering Shiny](https://mastering-shiny.org/) by Hadley Wickham
- [Quarto Shiny documentation](https://quarto.org/docs/interactive/shiny/)

### Code Comparison: Same Feature

::: {.panel-tabset}

#### Observable JS

```{ojs}
//| eval: false
//| echo: true
// Reactive filter with visualization
viewof species = Inputs.select(
  ["All", "Adelie", "Chinstrap", "Gentoo"],
  {label: "Species"}
)

filtered = species === "All"
  ? data
  : data.filter(d => d.species === species)

Plot.plot({
  marks: [
    Plot.dot(filtered, {x: "bill_length", y: "bill_depth", fill: "species"})
  ]
})
```

#### Shiny

```{r}
#| eval: false
#| echo: true
# UI
selectInput("species", "Species",
            choices = c("All", "Adelie", "Chinstrap", "Gentoo"))
plotOutput("scatter")

# Server
filtered <- reactive({
  if (input$species == "All") data
  else data %>% filter(species == input$species)
})

output$scatter <- renderPlot({
  ggplot(filtered(), aes(bill_length, bill_depth, color = species)) +
    geom_point()
})
```

:::

## Security and Compliance

### Data Privacy Considerations

| Concern | Observable JS | Shiny |
|---------|--------------|-------|
| Data exposure | In browser (DevTools) | Stays on server |
| PHI/PII | Needs de-identification | Server-side processing |
| HIPAA | Challenging | Feasible |
| Audit logging | Manual | Built-in (Connect) |

### Observable JS Security Model

**Risks:**

- All data sent to the browser is accessible to users
- Cannot truly hide business logic
- API keys should never be embedded

**Mitigations:**

- Pre-aggregate sensitive data
- Use server APIs for sensitive queries
- Embed only de-identified data

### Shiny Security Model

**Advantages:**

- Data never leaves the server
- Can integrate with enterprise authentication (LDAP, SAML)
- Audit trails available on Posit Connect
- Row-level security through user context

**Requirements:**

- Secure server configuration
- HTTPS enforcement
- Session timeout policies
- Regular security updates

### Regulatory Compliance

For FDA-regulated environments (21 CFR Part 11) or similar:

| Requirement | Observable JS | Shiny |
|-------------|--------------|-------|
| E-signatures | N/A | Connect supports |
| Audit trail | Limited | Full support |
| Access controls | Basic | Role-based |
| Validation | Challenging | Established |

::: {.callout-warning}
## Regulatory Note
For GxP environments, Shiny with Posit Connect provides established validation pathways. Observable JS may require additional validation effort.
:::

## Use Case Recommendations

### When to Choose Observable JS

::: {.callout-tip}
## Best Fit Scenarios

1. **Self-contained deliverables**: Client reports, manuscripts, presentations
2. **Wide distribution**: Reports going to many recipients
3. **Limited IT support**: No server infrastructure available
4. **Exploratory visualization**: Focus on data exploration over complex statistics
5. **Offline access needed**: Field work, conferences, travel
6. **Simple interactivity**: Filters, tooltips, basic calculations
:::

### When to Choose Shiny

::: {.callout-tip}
## Best Fit Scenarios

1. **Complex statistical methods**: Mixed models, Bayesian analysis, survival curves
2. **Large datasets**: More than ~50 MB of data
3. **Database integration**: Live queries, real-time data
4. **Sensitive data**: PHI, PII, proprietary information
5. **Regulated environments**: Pharma, finance, healthcare
6. **Enterprise deployment**: Integration with existing R infrastructure
:::

### Hybrid Approaches

For some projects, combining both technologies may be optimal:

```{mermaid}
%%| fig-cap: "Hybrid Architecture"
%%| fig-width: 5
%%| fig-height: 2.5
flowchart LR
    subgraph Backend
        A[R] --> B[JSON]
    end
    B --> C
    subgraph Report
        C[OJS] --> D[HTML]
    end
```

**Hybrid workflow:**

1. Use R for complex statistical analysis
2. Export results to JSON or Parquet
3. Use Observable JS for interactive visualization
4. Distribute as static HTML

## Conclusion

### Summary Comparison

| Dimension | Observable JS | Shiny |
|-----------|--------------|-------|
| Best for | Distribution, visualization | Statistics, security |
| Deployment | Static hosting | Server infrastructure |
| Data size limit | ~50 MB practical | Limited by server |
| Statistical depth | Basic | Comprehensive |
| Learning curve (R users) | Moderate | Low |
| Offline capability | Yes | No |
| Enterprise features | Limited | Extensive |

### Decision Framework

```{mermaid}
%%| fig-cap: "Technology Selection Flowchart"
%%| fig-width: 6
%%| fig-height: 3.5
flowchart LR
    A[New Report] --> B{Complex?}
    B -->|Yes| C[Shiny]
    B -->|No| D{Big data?}
    D -->|Yes| C
    D -->|No| E{Sensitive?}
    E -->|Yes| C
    E -->|No| F[Observable]
```

### Final Recommendations

1. **Start with your constraints**: Server availability, data sensitivity, and statistical requirements often dictate the choice.

2. **Consider your audience**: Technical users may appreciate Shiny's depth; broader audiences benefit from Observable's portability.

3. **Invest in learning both**: The skills complement each other, and hybrid approaches are increasingly common.

4. **Prototype quickly**: Both technologies allow rapid prototyping within Quarto---try both before committing.

## Appendix: Quick Reference

### Observable JS Cheat Sheet

```javascript
// Inputs
viewof x = Inputs.range([0, 100], {label: "Value"})
viewof y = Inputs.select(["A", "B", "C"], {label: "Category"})
viewof z = Inputs.checkbox(["X", "Y", "Z"], {label: "Options"})

// Data
data = FileAttachment("file.csv").csv({typed: true})

// Filtering
filtered = data.filter(d => d.value > x && y.includes(d.category))

// Plotting
Plot.plot({
  marks: [
    Plot.dot(filtered, {x: "xvar", y: "yvar", fill: "group"})
  ]
})
```

### Shiny Cheat Sheet

```r
# UI Inputs
sliderInput("x", "Value", 0, 100, 50)
selectInput("y", "Category", c("A", "B", "C"))
checkboxGroupInput("z", "Options", c("X", "Y", "Z"))

# Server
filtered <- reactive({
  data %>% filter(value > input$x, category %in% input$y)
})

output$plot <- renderPlot({
  ggplot(filtered(), aes(xvar, yvar, color = group)) + geom_point()
})
```

## References

- Wickham, H. (2021). *Mastering Shiny*. O'Reilly Media. <https://mastering-shiny.org/>
- Quarto Team. (2024). *Quarto Documentation*. Posit. <https://quarto.org/>
- Observable, Inc. (2024). *Observable Plot Documentation*. <https://observablehq.com/plot/>
- Posit. (2024). *Shiny Documentation*. <https://shiny.posit.co/>
